---
title: "STA363: Project"
author: "Sihang Zhao, Haoxuan Sun"
date: "2026-02-19"
output:
  pdf_document:
    toc: true
    toc_depth: 3
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.asp =.5)
```


```{r}
library(ggplot2)
library(rpart)
library(rattle)
library(rpart.plot)
library(nnet)
library(dplyr)
library(MASS)
library(tidyr)
library(naivebayes)
library(corrplot)
library(caret)
mydata<-read.csv("/Users/cosmozhao/Desktop/NBA Player Salaries (2022-23 Season) export 2026-02-09 18-59-01.csv")
```

\newpage
# EXECUTIVE SUMMARY AND CONCLUSION


\newpage
# AI STATEMENT
No generative AI tools were used in the process of conducting the analysis, performing the statistical modeling, or writing the substantive content of this report. All data analysis, coding, interpretation, and written explanations were completed independently by me, Sihang Zhao, as the sole author of this work. The only digital tool used during the writing process was Grammarly, and it was used exclusively for minor grammar and spelling corrections.


\newpage
# Project Title: Linguistic Traits and Authorship Prediction in Political Speeches of the United States
# INTRODUCTION


# ANALYSIS PROCEDURES


\newpage
\vspace*{\fill}
\begin{center}
\textbf{RESULTS ARE SHOWN BELOW}
\end{center}
\vspace*{\fill}


\newpage
# Section 1: EDA
