---
title: "STA363: Project"
author: "Sihang Zhao, Haoxuan Sun"
date: "2026-02-19"
output:
  pdf_document:
    toc: true
    toc_depth: 3
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message= FALSE, fig.asp = .6)
```


```{r}
library(ggplot2)
library(rpart)
library(rattle)
library(rpart.plot)
library(nnet)
library(dplyr)
library(MASS)
library(tidyr)
library(naivebayes)
library(corrplot)
library(caret)
library(knitr)
library(kableExtra)

mydata<-read.csv('/Users/cosmozhao/Desktop/NBA Player Salaries (2022-23 Season).csv')


```

\newpage
# EXECUTIVE SUMMARY AND CONCLUSION


\newpage
# AI STATEMENT
No generative AI tools were used in the process of conducting the analysis, performing the statistical modeling, or writing the substantive content of this report. All data analysis, coding, interpretation, and written explanations were completed independently by me, Sihang Zhao, as the sole author of this work. The only digital tool used during the writing process was Grammarly, and it was used exclusively for minor grammar and spelling corrections.


\newpage
# Project Title: 
# INTRODUCTION


# ANALYSIS PROCEDURES


\newpage
\vspace*{\fill}
\begin{center}
\textbf{RESULTS ARE SHOWN BELOW}
\end{center}
\vspace*{\fill}


\newpage
# Section 1: Data and Goal Overview

For this analysis, we use a dataset containing NBA player statistics and salaries from the 2022–2023 season. The dataset consists of (467) observations and (52) variables, where each observation represents an individual NBA player. The response variable is player salary, measured in U.S. dollars. While the dataset includes categorical variables such as player name, team, and position, these variables are excluded from the modeling process. Only quantitative performance statistics are used as predictors.

The goal of this project is to predict a player’s salary based solely on measurable on-court performance metrics. By focusing on numerical variables, we aim to assess the extent to which statistical productivity explains differences in player compensation.

The primary features considered include age, games played (GP), games started (GS), minutes per game (MP), field goals (FG), field goal percentage (FG%), three-point statistics, rebounds, assists, steals, blocks, turnovers, points per game, and other continuous box score or advanced metrics available in the dataset. These variables capture scoring efficiency, playing time, and overall player contribution.

This project is particularly interesting because salary determination in professional sports reflects performance valuation within a competitive market. By modeling salary as a function of quantitative performance indicators, we can evaluate how strongly observable productivity translates into financial compensation.



# Section 2: EDA

## Raw Data Glimpse
```{r}
# build d inside this chunk (so knitting always has it)
d <- mydata %>%
  dplyr::select(-X) %>%                             # remove X if it exists
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))) %>%
  head(6)

key <- c("Player.Name", "Salary")   # keep these visible in both tables
p <- 10                              # columns per table (adjust)

# split columns into blocks of size p
col_blocks <- split(names(d), ceiling(seq_along(names(d)) / p))

# Part A
d %>%
  dplyr::select(any_of(key), all_of(setdiff(col_blocks[[1]], key))) %>%
  kable(booktabs = TRUE, caption = "Raw data preview (Part A)") %>%
  kable_styling(latex_options = "scale_down")

# Part B (only if exists)
if (length(col_blocks) > 1) {
  d %>%
    dplyr::select(any_of(key), all_of(setdiff(col_blocks[[2]], key))) %>%
    kable(booktabs = TRUE, caption = "Raw data preview (Continued)") %>%
    kable_styling(latex_options = "scale_down")
}
```

## Check for multicollinearity among features

# Attention about title and caption

```{r,fig.asp = 1}
# Select numeric variables only
num_data <- mydata %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select(-any_of("X"))   # remove index column if present

# Compute correlation matrix
cor_mat <- cor(num_data, use = "complete.obs")

# Plot correlogram
corrplot(cor_mat,
         method = "circle",
         type = "upper",
         order = "hclust",
         tl.col = "black",
         tl.cex = 0.6)

```

blahblahblahblahblahblah


## The Data 
```{r}
mydata_clean <- mydata %>%
  dplyr::select(
    Salary,
    Age,
    MP,       # choose this instead of PTS
    AST.,
    TRB,
    BPM,
    TOV.
  )

knitr::kable(mydata_clean[1:10,], caption = "The First 10 Rows of Our Data")
```

these are features we are gonna use, blahblahblah


```{r}
glossary <- data.frame(
  Acronym = c("Salary", "Age", "MP", "AST", "TRB", "BPM", "TOV"),
  Explanation = c(
    "Annual player salary (USD)",
    "Player age (years)",
    "Minutes played per game",
    "Assists per game",
    "Total rebounds per game",
    "Box Plus/Minus — overall impact metric",
    "Turnovers per game"
  )
)

kable(glossary,
      booktabs = TRUE,
      col.names = c("Acronym", "Explanation"),
      caption = "Table: Variable Definitions") %>%
  kable_styling(latex_options = "hold_position")
```

## Data Exploration




\newpage
# Appendix A: The Data

```{r}
knitr::kable(mydata_clean, title = "The Original Data")
```
